---
title: "SI_JHU_W3Quizz"
author: "Linda Angulo Lopez"
date: "11/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## StatisticalInference W3_Quizz

In a population of interest, a sample of 9 men yielded a sample average brain
volume of 1,100cc and a standard deviation of 30cc. What is a 95% Student's T
confidence interval for the mean brain volume in this new population?

we know that: 

`CI = mu + c(-1,1)*sigma*qt(0.975,df=n-1)/sqrt(n)`

```{r CI_t9MenBrain}

mu <- 1100
sigma <- 30
n <- 9
CI <- mu + c(-1,1)*sigma*qt(0.975,df=n-1)/sqrt(n)
CI
#[1] 1076.94 1123.06

```


A diet pill is given to 9 subjects over six weeks. The average difference in
weight (follow up - baseline) is -2 pounds. What would the standard deviation
of the difference in weight have to be for the upper endpoint of the 95% T
confidence interval to touch 0?

from, 
`CI = 0 = mu + c(-1,1)*sigma*qt(0.975,df=n-1)/sqrt(n)`
plug in, 
`CI = 0 = -2 + sigma*qt(0.975,8)/sqrt(n)`
so, 
`sigma = -mu*sqrt(n)/qt(0.975,df=n-1)`

```{r sd_DietPill_9_6wks}

mu <- -2
n <- 9
sigma <- -mu*sqrt(n)/qt(0.975,df=n-1)
sigma

```

In an effort to improve running performance, 5 runners were either given a
protein supplement or placebo. Then, after a suitable washout period, they 
were given the opposite treatment. Their mile times were recorded under both 
the treatment and placebo, yielding 10 measurements with 2 per subject. The
researchers intend to use a T test and interval to investigate the treatment.

Should they use a paired or independent group T test and interval?


```{r Paired_Independent}

#We want to know the impact of the substitution of treatment and
#placebo on the same person, from two independant groups, so
print("We should use a paired t-test.")
```


In a study of emergency room waiting times, investigators consider a new and
the standard triage systems. To test the systems, administrators selected 20
nights and randomly assigned the new triage system to be used on 10 nights and
the standard system on the remaining 10 nights. They calculated the nightly
median waiting time (MWT) to see a physician. 

The average MWT for the new system was 3 hours with a variance of 0.60 while
the average MWT for the old system was 5 hours with a variance of 0.68.

Consider the 95% confidence interval estimate for the differences of the mean 
MWT associated with the new system. Assume a constant variance. What is the
interval? Subtract in this order (New System - Old System).

We know, 
`S_p = sqrt(((n_new-1)*v_new +(n_old-1)*v_old)/(n_new + n_old-2))`
and we know, 
`CI = (mu_new - mu_old + c(-1,1)*qt(0.975,n_new+n_old-2)*S_p*sqrt(1/n_new + 1/n_old))`
so we can get S_p & t, 
then plug in, 
`R function, qt()`
to get the new CI.


```{r newCI_waitingTime}

mu_old <-5
mu_new <- 3
n_old <- 10
n_new <-  10
v_new <- 0.6
v_old<- 0.68
S_p <- sqrt(((n_new-1)*v_new +(n_old-1)*v_old)/(n_new + n_old-2))
CI <- (mu_new - mu_old + c(-1,1)*qt(0.975,n_new+n_old-2)*S_p*sqrt(1/n_new + 1/n_old))
CI
#[1] -2.751649 -1.248351
S_p

```



```{r newCI_waitingTime2 }
qt(0.975,n_new+n_old-2)

```



```{r newCI_waitingTime3}
CI
#[1] -2.751649 -1.248351

```


Suppose that you create a 95% T confidence interval. You then create a 90%
interval using the same data. What can be said about the 90% interval with
respect to the 95% interval?

```{r TIntervals95_90}
print("As tdf,975>tdf,95, the 90% interval will be narrower than the 95% interval.")

```


To further test the hospital triage system, administrators selected 200 nights
and randomly assigned a new triage system to be used on 100 nights and a
standard system on the remaining 100 nights. They calculated the nightly
median waiting time (MWT) to see a physician. The average MWT for the new
system was 4 hours with a standard deviation of 0.5 hours while the average
MWT for the old system was 6 hours with a standard deviation of 2 hours.

Consider the hypothesis of a decrease in the mean MWT associated with the new
treatment.


What does the 95% independent group confidence interval with unequal variances
suggest vis a vis this hypothesis? (Because there's so many observations per
group, just use the Z quantile instead of the T.)

we know, 
`S_p = sqrt(((n_x - 1) * S_x^2 + (n_y - 1) * S_y^2)/(n_x + n_y - 2))`
and we know, 
`CI =  mu_x - mu_y + c(-1, 1) * qnorm(quantile) * S_p * (1 / n_x + 1 / n_y)^.5`
so we can plug in:

```{r TwoSidedMWT_95CI}

quantile = 0.975
n_y <- 100     
n_x <- 100     
S_y <- 0.50    
S_x <- 2       
mu_y <- 4      
mu_x <- 6      


S_p <- sqrt(((n_x - 1) * S_x^2 + (n_y - 1) * S_y^2)/(n_x + n_y - 2))

CI <-  mu_x - mu_y + c(-1, 1) * qnorm(quantile) * S_p * (1 / n_x + 1 / n_y)^.5
round(CI,2)

#[1] 1.6 2.4

print("When subtracting (old - new) the interval is above zero. 
so the average meeting time has gone down, the new system appears
to be effective.")

```


Suppose that 18 obese subjects were randomized, 9 each, to a new diet pill and a
placebo. Subjects’ body mass indices (BMIs) were measured at a baseline and again
after having received the treatment or placebo for four weeks. The average
difference from follow-up to the baseline (followup - baseline) was −3 kg/m2 for the
treated group and 1 kg/m2 for the placebo group. 

The corresponding standard deviations of the differences was 1.5 kg/m2 for the 
treatment group and 1.8 kg/m2 for the placebo group. 

Does the change in BMI over the four week period appear to differ between the
treated and placebo groups? Assuming normality of the underlying data and a common
population variance, calculate the relevant *90%* t confidence interval. Subtract in
the order of (Treated - Placebo) with the smaller (more negative) number first.

calculate, 
`S_p = sqrt(((n_x - 1) * S_x^2 + (n_y - 1) * S_y^2)/(n_x + n_y - 2))`

plug in S_p, canculate, 
`CI <-  mu_y - mu_x + c(-1, 1) * qt(quantile, df=n_y+n_x-2) * S_p * (1 / n_x + 1 / n_y)^.5`

```{r cars}

quantile = 0.95 

n_y <- 9 
n_x <- 9 
S_y <- 1.5
S_x <- 1.8
mu_y <- -3
mu_x <- 1


S_p <- sqrt(((n_x - 1) * S_x^2 + (n_y - 1) * S_y^2)/(n_x + n_y - 2))

CI <-  mu_y - mu_x + c(-1, 1) * qt(quantile, df=n_y+n_x-2) * S_p * (1 / n_x + 1 / n_y)^.5
round(CI,3)
CI
#[1] -5.364 -2.636


```



